Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
Edits

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.~a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program with a fixed amount of
storage terminates.  Since such programs are necessarily finite state, our
problem is much easier than Turing's, but is more relevant to analysing computer
programs.

@@

@@

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program, including
nested loops.  We can use this to analyse all the loops at once,
or one at a time.
% In contrast to previous approaches to program synthesis, we
% found that using genetic programming was very effective at generating
% candidate termination and non-termination proofs.

@@

\item Our solver is based on a series of source-to-source
transformations.  As well as leading to a concise implementation, this has
the side effect of ensuring that our proofs are correct for the exact binary
produced by a particular compiler, rather than for an abstraction of the
program.

@@

The problem of program verification can be reduced to the problem of finding
solutions to a second-order
constraint~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
Our intention is to apply this approach to termination analysis.  In this
section we show how several variations of both the termination and the
non-termination problem can be defined in second-order logic.

@@

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.
%
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
that all quantified variables range over vectors.

@@

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Our technique proves the termination of each loop separately, 
with the exception of nested loops, whose termination we encode in the same formula. For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $T_o$  over-approximates $L_1$'s transition relation.
\item $S$ is a summary of $L_2$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.
\end{itemize}

@@

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
   S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
   & R_2(i,j,n) > 0 \, \wedge \\
   & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
   i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
   i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
   & T_o(i'+1, j', n') \, \wedge \\
   & R_1(i, j, n) > 0 \, \wedge \\
   & R_1(i, j, n) > R_1(i'+1, j', n')
\end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

@@

As well as efficiency, compiling the original source code of the program
under analysis has the advantage of guaranteeing that we are verifying the
exact semantics of the program as understood by a compiler, rather than some
ad-hoc abstraction of its semantics.

@@

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula {\bf [GT]} for any loop $L$ is satisfiable
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.  Corollary~\ref{cor:cegis-newc}
 shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

@@

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover,
named {\sc Juggernaut}, on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of 180\,s, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

@@

>>>>

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.~a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program with a fixed amount of
storage terminates.  Since such programs are necessarily finite state, our
problem is much easier than Turing's, but is a better fit for analysing computer
programs.

@@

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program, including
nested loops.  We can use this to analyse all the loops at once,
or one at a time.  Our treatment handles termination and non-termination uniformly:
both properties are captured in the same second-order formula.

@@

@@

\item Our solver is based on a series of source-to-source
transformations.  As well as leading to a concise implementation, this
allows us to leverage existing, mature verification tools.

@@

The problem of program verification can be reduced to the problem of finding
solutions to a second-order
constraint~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
Our intention is to apply this approach to termination analysis.  In this
section we show how several variations of both the termination and the
non-termination problem can be uniformly defined in second-order logic.

@@

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.
%
\end{definition}
%
We note that by existentially quantifying over Skolem functions, formulae with arbitrary
first-order quantification can be brought into the synthesis fragment~\cite{hol-book}, so the
fragment is semantically less restrictive than it looks.
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the vector notation and write $x$ for $\vec{x}$, with the understanding
that all quantified variables range over vectors.

@@

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Full details of this construction appear in the extended version of this paper~\cite{extended-version}.
For the current example, the termination formula is given on the right side of 
Figure~\ref{fig:environment-model}:
%% Following the same design, let us assume that we have already proven the termination of loop $L_1$, and we are now trying to prove that $L_2$ and $L_3$ terminate, which we encode
%% as the conjunction of constraints on the right side of the figure:
$T_o$ is a summary of $L_1$ that over-approximates its transition relation;
$R_1$ and $R_2$ are ranking functions for $L_1$ and $L_2$, respectively.

@@

  i = i + 1;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists T_o, R_1, R_2 . \forall i, j, n, i', j', n' . \\
   i < n & \rightarrow T_o(\langle i, j, n \rangle , \langle i, 0, n \rangle) ~ \wedge \\
   j \leq i \wedge T_o(\langle i', j', n' \rangle, \langle i, j, n\rangle ) & \rightarrow R_2(i, j, n) > 0 ~ \wedge \\
     & R_2(i, j, n) > R_2(i, j+1, n) ~ \wedge \\
     & T_o(\langle i', j', n' \rangle, \langle i, j+1, n \rangle) ~ \wedge \\
   i < n \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow R_1(i, j, n) > 0 ~ \wedge \\
                                         & R_1(i, j, n) > R_1(i+1, j, n)
   \end{align*}
% \begin{align*}
%  \exists T_o, S, R_1, R_2 . \forall i, j, n, i', j', n' . \\
%    %% x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
%    %% x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
%    %% x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x, y, z) \, \wedge \\
%    S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' \leq i' & \rightarrow S(\langle i, j, n \rangle, \langle i', j'+1, n' \rangle) \, \wedge \\
%    & R_2(i,j,n) > 0 \, \wedge \\
%    & R_2(i,j,n) > R_2(i',j'+1,n') \, \wedge\\
%    i < n \wedge T_o(i, j, n) & \rightarrow S(\langle i, j, n \rangle, \langle i, 0, n \rangle) \, \wedge \\
%    i < n \wedge T_o(i, j, n) \wedge S(\langle i, j, n \rangle, \langle i', j', n' \rangle) \wedge j' > i' & \rightarrow \\
%    & T_o(i'+1, j', n') \, \wedge \\
%    & R_1(i, j, n) > 0 \, \wedge \\
%    & R_1(i, j, n) > R_1(i'+1, j', n')
% \end{align*}
\end{minipage}
\end{framed}
\caption{A program with nested loops and its termination formula\label{fig:environment-model}}
\end{figure*}

@@

% As well as efficiency, compiling the original source code of the program
% under analysis has the advantage of guaranteeing that we are verifying the
% exact semantics of the program as understood by a compiler, rather than some
% ad-hoc abstraction of its semantics.

@@

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula {\bf [GT]} for any loop $L$ is valid
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is valid, and hence satisfiable.  Corollary~\ref{cor:cegis-newc}
 shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

@@

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover,
named {\sc Juggernaut}, on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}\footnote{
The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of 180\,s, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

@@

Overall, we notice that the GP component was always the fastest to find
candidate solutions, {\sc ExplicitSearch} was always fastest
to find counterexample inputs, and {\sc CBMC} was always fastest at
proving correctness.  Each of these search methods has its strengths,
and removing any one would have had a significant negative impact on
overall performance.  Approximately 90\% of the total
runtime was spent in synthesising new candidate solutions, which matches
our expectations -- the (non-)termination proofs are difficult to
find, but easy to check.
